Processing genome scale tabular data with wormtable

### 2013-08-21 docs/paper/paper.tex
modified up to seek performance section

%\emph{The Results and Discussion may be combined into a single section or
%presented separately. They may also be broken into subsections with
%short, informative headings. In any case what should be described is the
%functionality of the software together with data on how its performance
%and functionality compare with and improve on functionally similar
%existing software. There should then be a discussion of the intended
%use of the software, and the benefits that are envisioned together,
%if possible, with an outline for the planned future development of
%new features.}
%
\subsection*{API}
The principle goal of wormtable is to provide efficient access to data 
using an easy to learn interface. 
The \texttt{Table} class provides the main interface,
and implements the standard Python sequence protocols. Efficient 
iteration over rows is provided by the \texttt{cursor} method, 
which takes as an argument the list of columns to read.
Only the values for the columns 
of interest are then retrieved, leading to considerable
time savings. The \texttt{cursor} function takes two additional 
arguments, \texttt{start} and \texttt{stop}, which specify 
rows of interest. This allows us to efficiently seek to an 
arbitrary location in the table and to read a given number 
of rows sequentially from this point.

@@

The API is straightforward, but it is flexible, powerful and 
extensible. This simplicity ensures that 
programmers of all experience levels  
can take full advantage of the powerful data 
processing facilities that wormtable provides. In the following subsections 
we illustrate the performance advantages of wormtable via some 
examples. These are not intended to be definitive benchmarks
and are simply concrete examples to demonstrate the type of improvements 
that can be expected by using wormtable over existing methods.

\subsection*{Scan performance}
The Variant Call Format (VCF) encodes information about variant sites in a genome 
as tab-delimited rows in a text file~\cite{da+11}.  
To illustrate the advantages of wormtable when performing calculations over 
a whole table, we converted a large 
publicly available VCF file (produced as part of the
Drosophila genetic reference panel~\cite{mr+12}) to wormtable format.
This VCF\cite{freeze.vcf}
consists of 15GB of uncompressed text and contains data from the 
whole genome over 6,146,611 rows.
Using the included \texttt{vcf2wt} program
the VCF was converted to wormtable format 
on a workstation with an Intel Xeon processor, 12GB of RAM 
and a single hard disk. The
conversion required approximately 69 minutes, and the size of the resulting  
wormtable was 10GB (using the smallest type required
to represent the data in each column).

>>>>

%\emph{The Results and Discussion may be combined into a single section or
%presented separately. They may also be broken into subsections with
%short, informative headings. In any case what should be described is the
%functionality of the software together with data on how its performance
%and functionality compare with and improve on functionally similar
%existing software. There should then be a discussion of the intended
%use of the software, and the benefits that are envisioned together,
%if possible, with an outline for the planned future development of
%new features.}
%
\subsection*{API}
The principle goal of wormtable is to provide efficient access to data 
using an easy to learn interface. 
In Python, the \texttt{Table} class provides the main interface,
and implements the standard Python sequence protocols. Efficient 
iteration over rows is provided by the \texttt{cursor} method, 
which takes as an argument the list of columns to read.
Only the values for the columns 
of interest are then retrieved, leading to considerable
time savings. The \texttt{cursor} function can take two additional 
arguments, \texttt{start} and \texttt{stop}, which specify 
the rows of interest. This allows us to efficiently seek to an 
arbitrary location in the table and to read a given number 
of rows sequentially from this point.

@@

The API is straightforward, but it is flexible, powerful and 
extensible. This simplicity ensures that 
programmers of all experience levels  
can take full advantage of the powerful data 
processing facilities that wormtable provides. In the following subsections 
we illustrate the performance advantages of wormtable via some 
examples. These are not intended to be definitive benchmarks
but are simple examples to demonstrate the type of improvements 
that can be expected by using wormtable over existing methods.

\subsection*{Scan performance}
The Variant Call Format (VCF) encodes information about variant sites in a genome 
as tab-delimited rows in a text file~\cite{da+11}. The VCF format is one of 
the most commonly used formats to store genomic data from next generation sequencing.
To illustrate the advantages of wormtable when performing calculations over 
a whole table, we converted a large 
publicly available VCF file (produced as part of the
Drosophila genetic reference panel~\cite{mr+12}) to wormtable format.
This VCF\cite{freeze.vcf}
consists of 15GB of uncompressed text and contains data from the 
whole genome over 6,146,611 rows.
Using the included \texttt{vcf2wt} program
the VCF was converted to wormtable format 
on a workstation with an Intel Xeon processor, 12GB of RAM 
and a single hard disk. The
conversion required approximately 69 minutes, and the size of the resulting  
wormtable was 10GB (using the smallest type required
to represent the data in each column).
